Exploring the 2012 PISA data
========================================================

Data sources: http://www.oecd.org/pisa/pisaproducts/datavisualizationcontest.htm

This is the code that will produce the plots in Di Cook's Les Diablerets seminar on Exploratory Data Analysis. 

Load all the libraries that will generally be needed.

```{r load_libraries, message=FALSE}
library(stringr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(lubridate)
library(maps)
library(ggmap)
library(rworldmap)
library (grid)    
library (scales) 
library(GGally)
```

Read in the data. You will need to get a copy of the .rda files from the OECD PISA web site. These files are large.

```{r load_data, echo=FALSE, cache=FALSE}
#setwd("PISA")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name){
 
  template <- c("2012.rda", "2012dict.rda")
 
  file_name %>% 
    vapply(str_join, template, template) %>% 
    file.path(".", "data", .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# clean
rm(fn_build, sets)
```

A little bit of convenience code for being able to sort through the variable definitions.

```{r modify_dictionaries, echo=FALSE, cache=FALSE}
# function to convert to data-frames
fn_make_df <- function(named_vector){
  data.frame(
    variable = attr(named_vector, "names"),
    description = named_vector,
    row.names = NULL
  )
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict) 
dict_parent2012 <- fn_make_df(parent2012dict) 
dict_school2012 <- fn_make_df(school2012dict) 
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
dict_student2012 <- fn_make_df(student2012dict) 

# clean
rm(fn_make_df)
rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

Working with the maps, convenience functions.

```{r usefulfn, echo=FALSE, cache=FALSE}
extractPolygons <- function(shapes) {

  dframe <- ldply(1:length(shapes@polygons), function(i) {
    ob <- shapes@polygons[[i]]@Polygons
    dframe <- ldply(1:length(ob), function(j) {
      x <- ob[[j]]
      co <- x@coords
      data.frame(co, order=1:nrow(co), group=j)
    })
    dframe$region <- i
    dframe$name <- shapes@polygons[[i]]@ID
    dframe
  })
  # construct a group variable from both group and polygon:
  dframe$group <- interaction(dframe$region, dframe$group)
  
  dframe
}

# To get a blank background on map
new_theme_empty <- theme_bw()
new_theme_empty$line <- element_blank()
new_theme_empty$rect <- element_blank()
new_theme_empty$strip.text <- element_blank()
new_theme_empty$axis.text <- element_blank()
new_theme_empty$plot.title <- element_blank()
new_theme_empty$axis.title <- element_blank()
new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
```

Take a quick browse through the size of the data to see what we have.

```{r quicklook, fig.width=8, fig.height=8, warning=FALSE, message=FALSE, cache=FALSE}
head(dict_student2012, 20)
sort(table(student2012$CNT))
unique(student2012$CNT)
tail(dict_student2012, 5)
x <- student2012[,c("CNT", "PV1MATH", "PV2MATH", "SENWGT_STU")]
x$PV1MATH <- as.numeric(x$PV1MATH)
x$PV2MATH <- as.numeric(x$PV2MATH)
x$SENWGT_STU <- as.numeric(x$SENWGT_STU)
morder <- summarise(group_by(x, CNT), mave=weighted.mean(PV1MATH, w=SENWGT_STU, na.rm=T))
x$CNT <- factor(x$CNT, levels=morder$CNT[order(morder$mave)])
qplot(PV1MATH, PV2MATH, data=x) + facet_wrap(~CNT) + theme(aspect.ratio=1)
x <- student2012[student2012$CNT=="Switzerland",c("PV1MATH", "PV2MATH","PV3MATH","PV4MATH","PV5MATH")]
ggpairs(x)
```

```{r cntnames, warning=FALSE, message=FALSE, cache=FALSE}
colnames(student2012)[1] <- "name"
#school2012$name <- as.character(school2012$name)
#student2012$name <- as.character(student2012$name)

fix_country = function(df) {
df$name = as.character(df$name)
df$name[df$name=="Serbia"] <- "Republic of Serbia"
df$name[df$name=="Korea"] <- "South Korea"
df$name[df$name=="Chinese Taipei"] <- "Taiwan"
df$name[df$name=="Slovak Republic"] <- "Slovakia"
df$name[df$name=="Russian Federation"] <- "Russia"
df$name[df$name=="Perm(Russian Federation)"] <- "Russia"
df$name[df$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
df$name[df$name=="China-Shanghai"] <- "China"
df$name[df$name=="China-Macau"] <- "China"
df$name[df$name=="Connecticut (USA)"] <- "United States of America"
df$name[df$name=="Florida (USA)"] <- "United States of America"
df$name[df$name=="Massachusetts (USA)"] <- "United States of America"

df
}

shorten_countrynames = function(df) {
df$name = as.character(df$name)
df$name[df$name=="Hong Kong S.A.R."] <- "Hong Kong"
df$name[df$name=="Republic of Serbia"] <- "Serbia"
df$name[df$name=="United Arab Emirates"] <- "UAE"
df$name[df$name=="United States of America"] <- "USA"
df$name[df$name=="United Kingdom"] <- "UK"

df
}
student2012 = fix_country(student2012)
student2012 = shorten_countrynames(student2012)
```

Calculate summaries by gender.

```{r dataprep, fig.width=8, fig.height=8, warning=FALSE, message=FALSE, cache=FALSE}
student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 551, 634)]
colnames(student2012.sub)[1] <- "name"
student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
student2012.sub$W_FSTUWT <- as.numeric(student2012.sub$W_FSTUWT)
# Set up country names to match map
#student2012.sub$name <- as.character(student2012.sub$name)
#student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
#student2012.sub$name[student2012.sub$name=="Macao-China"] <- "China"
#student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
#student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
#student2012.sub$name[student2012.sub$name=="Hong Kong S.A.R."] <- "Hong Kong"
#student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
#student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
#student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
#student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
#student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
#student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
#student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
#student2012.sub$name[student2012.sub$name=="United States of America"] <- "USA"

student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                  w2mathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=W_FSTUWT[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=W_FSTUWT[ST04Q01=="Female"], na.rm=T),
                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
                  wreadgap=weighted.mean(PV1READ[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1READ[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
                  wsciencegap=weighted.mean(PV1SCIE[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1SCIE[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean Difference", ylab="Weighted Mean Difference",
      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
summary(student2012.sub.summary.gap$mathgap-student2012.sub.summary.gap$wmathgap)
qplot(wmathgap, w2mathgap, data=student2012.sub.summary.gap, xlab="Senate Weighted Mean Difference", ylab="Final Student Weighted Mean",
      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)

# Rough grouping by p-values
student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"

# Extract map polygons for modern world
world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
# Check
head(world.polys)
sort(unique(world.polys$name))
sort(unique(as.character(student2012$CNT)))
sort(unique(as.character(student2012$CNT)))[!(sort(unique(as.character(student2012$CNT))) %in% sort(unique(world.polys$name)))]
# Now add data
#student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Serbia"] <- "Republic of Serbia"
world.polys$name[world.polys$name=="Republic of Serbia"] <- "Serbia"
world.polys$name[world.polys$name=="United States of America"] <- "USA"
world.polys$name[world.polys$name=="United Arab Emirates"] <- "UAE"
world.polys$name[world.polys$name=="United Kingdom"] <- "UK"
world.polys$name[world.polys$name=="Hong Kong S.A.R."] <- "Hong Kong"
student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
```

```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, cache=FALSE}
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
qplot(name, mathgap, data=student2012.sub.summary.gap, size=I(5), color=msig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5), limits=c(-30,30)) + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
# Test colors
#qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
#  xlab("") +
#  scale_colour_manual("Significant", values=c("male"=dichromat("skyblue"), "female"=dichromat("pink"), "none"=dichromat("lightgreen"))) +
#  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
#  scale_size("Prop male") +
#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
#ggsave("gendermathgap.pdf", width=7, height=14)
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue"), size=I(5)) + 
  xlab("") +
  geom_point(aes(y=maxfemale), colour="pink", size=I(5)) + 
  scale_y_continuous("High Math") + coord_flip() + theme_bw()
#ggsave("gendermathtop.pdf", width=3.5, height=7)
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue"), size=I(5)) + 
  xlab("") +
  geom_point(aes(y=minfemale), colour="pink", size=I(5)) + 
  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
#ggsave("gendermathbottom.pdf", width=3.5, height=7)
#
```

```{r mathmap, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, cache=FALSE}
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
```

```{r genderread, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, cache=FALSE}
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
qplot(name, readgap, data=student2012.sub.summary.gap, color=rsig, size=I(5)) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Reading Score Gap") + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
#ggsave("genderreadgap.pdf", width=5, height=8)
```

```{r readmap, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, cache=FALSE}
#ggsave("gendermathmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
```

```{r genderscience, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, cache=FALSE}
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
qplot(name, sciencegap, data=student2012.sub.summary.gap, color=ssig, size=I(5)) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Science Score Gap") + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
#ggsave("gendergap.pdf", width=10, height=14)
````

```{r maps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, cache=FALSE}
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
```

```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE, cache=FALSE}
student.luke <- student2012[,c("CNT","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
colnames(student.luke)[1] <- "name"
#study.luke <- melt(student.luke,id=c("name","ST57Q01"))
studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH)) + xlab("Hours spent out of school studying per week") + ylab("Math score") + facet_wrap(~name) + stat_smooth(se=F)
studyplot
```


```{r studyingbar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
studybar <- ggplot(data = student.luke, aes(ST57Q01) ) + geom_histogram(binwidth=5) + xlab("Hours spent studying")
studybar
```

```{r truant, echo=FALSE, message = FALSE, warning = FALSE, cache=FALSE}
#student2012$CNT = student2012$name
student2012$OECD <- as.character(student2012$OECD)
#student2012[student2012$name == "Albania", ]$OECD = "Albania"
student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])

student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
colnames(student2012.sub)[1] <- "name"

for (i in 13:26) {
    student2012.sub[, i] <- as.character(student2012.sub[, i])
    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
    student2012.sub[student2012.sub[, i] == "Yes", i] <- "1"
    student2012.sub[student2012.sub[, i] == "No", i] <- "0"
    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
}
for (i in 27:31) {
    student2012.sub[, i] <- as.character(student2012.sub[, i])
    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
    student2012.sub[student2012.sub[, i] == "None", i] <- "0"
    student2012.sub[student2012.sub[, i] == "One", i] <- "1"
    student2012.sub[student2012.sub[, i] == "Two", i] <- "2"
    student2012.sub[student2012.sub[, i] == "Three or more", i] <- "3"
    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
}
student2012.sub[, 32] <- as.character(student2012.sub[, 32])
student2012.sub[is.na(student2012.sub[,32]), 32] <- ""
student2012.sub[student2012.sub[, 32] == "0-10 books ", 32] <- "0"
student2012.sub[student2012.sub[, 32] == "11-25 books ", 32] <- "1"
student2012.sub[student2012.sub[, 32] == "26-100 books ", 32] <- "2"
student2012.sub[student2012.sub[, 32] == "101-200 books ", 32] <- "3"
student2012.sub[student2012.sub[, 32] == "201-500 books ", 32] <- "4"
student2012.sub[student2012.sub[, 32] == "More than 500 books", 32] <- "5"
student2012.sub[, 32] <- as.numeric(student2012.sub[, 32])

#aggregate possessions
student2012.sub$numposs <- apply(student2012.sub[, 13:26], 1, sum, na.rm = T)
student2012.sub$numedposs <- apply(student2012.sub[, 27:31], 1, sum, na.rm = T)
```

```{r par_athomebar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE, cache=FALSE}
athome = student2012[,c("CNT", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
colnames(athome)[1] <- "name"
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$parents <- factor(athome$parents, levels=c("Both", "Mother", "Father", "Neither"))
#athome$name <- reorder(athome$name,athome$PV1MATH,mean, decreasing=F)
pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
pbarhome
```

```{r par_athome, fig.width=12, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE, cache=FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N")) 
phome
```

```{r poss, echo=FALSE, message = FALSE, warning = FALSE, cache=FALSE}
#student2012$CNT = student2012$name
student2012$OECD <- as.character(student2012$OECD)
#student2012[student2012$name == "Albania", ]$OECD = "Albania"
student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])

student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
colnames(student2012.sub)[1] <- "name"

for (i in 13:26) {
    student2012.sub[, i] <- as.character(student2012.sub[, i])
    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
    student2012.sub[student2012.sub[, i] == "Yes", i] <- "1"
    student2012.sub[student2012.sub[, i] == "No", i] <- "0"
    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
}
for (i in 27:31) {
    student2012.sub[, i] <- as.character(student2012.sub[, i])
    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
    student2012.sub[student2012.sub[, i] == "None", i] <- "0"
    student2012.sub[student2012.sub[, i] == "One", i] <- "1"
    student2012.sub[student2012.sub[, i] == "Two", i] <- "2"
    student2012.sub[student2012.sub[, i] == "Three or more", i] <- "3"
    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
}
student2012.sub[, 32] <- as.character(student2012.sub[, 32])
student2012.sub[is.na(student2012.sub[,32]), 32] <- ""
student2012.sub[student2012.sub[, 32] == "0-10 books ", 32] <- "0"
student2012.sub[student2012.sub[, 32] == "11-25 books ", 32] <- "1"
student2012.sub[student2012.sub[, 32] == "26-100 books ", 32] <- "2"
student2012.sub[student2012.sub[, 32] == "101-200 books ", 32] <- "3"
student2012.sub[student2012.sub[, 32] == "201-500 books ", 32] <- "4"
student2012.sub[student2012.sub[, 32] == "More than 500 books", 32] <- "5"
student2012.sub[, 32] <- as.numeric(student2012.sub[, 32])

#aggregate possessions
student2012.sub$numposs <- apply(student2012.sub[, 13:26], 1, sum, na.rm = T)
student2012.sub$numedposs <- apply(student2012.sub[, 27:31], 1, sum, na.rm = T)
```

```{r albania-possessions, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE, cache=FALSE}
#df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("name", "numposs", "OECD"))
df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math)), id=c("name", "numposs", "OECD"))

#df$group = paste0(df$name, df$variable)
#qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
qplot(x = numposs, y = value, data = df, facets = ~name, geom = "line") + ylab("Math score") + xlab("Number of possessions") + theme_bw()
```

```{r}
# Internet use
student2012dict[307:323]
nm <- unlist(strsplit(student2012dict[307:323]," - "))[seq(2, 34, 2)]
ch.internet <- student2012[student2012$CNT=="Switzerland",307:323]
colnames(ch.internet) <- nm
sort(apply(ch.internet, 2, mean, na.rm=T))
recode.manually <- function(x){
  x <- as.character(x)
  x[x=="Never or hardly ever"] <- 0
  x[x=="Once or twice a month"] <- 1
  x[x=="Once or twice a week"] <- 2
  x[x=="Almost every day"] <- 3
  x[x=="Every day"] <- 4
  as.numeric(x)
}
for (i in 1:ncol(ch.internet)) 
  ch.internet[,i] <- recode.manually(ch.internet[,i])
ch.internet.m <- melt(ch.internet)
ch.internet.m.ave <- summarise(group_by(ch.internet.m, variable), av = mean(value, na.rm=T))
ch.internet.m.ave$variable <- factor(ch.internet.m.ave$variable, levels=ch.internet.m.ave$variable[order(ch.internet.m.ave$av)])
qplot(variable, av, data=ch.internet.m.ave, size=I(5), ylab="Average usage", xlab="", ylim=c(0,4)) + coord_flip()
```

```{r}
student2012.sub <- student2012[, c(1, 62, 501)]
colnames(student2012.sub)[2] <- "TVs"
colnames(student2012.sub)[3] <- "math"
student2012.sub <- student2012.sub[!is.na(student2012.sub[,2]),]
student2012.sub[, 2] <- as.character(student2012.sub[, 2])
student2012.sub[student2012.sub[, 2] == "None", 2] <- 0
student2012.sub[student2012.sub[, 2] == "One", 2] <- 1
student2012.sub[student2012.sub[, 2] == "Two", 2] <- 2
student2012.sub[student2012.sub[, 2] == "Three or more", 2] <- 3
student2012.sub[, 2] <- as.numeric(student2012.sub[, 2])
df <- student2012.sub %>% group_by(name, TVs) %>% summarise(mmath = mean(math))
df2 <- df %>% group_by(name) %>% summarise(diff = mmath[TVs==0]-mmath[TVs==3])
df$name <- factor(df$name, levels=unique(df$name)[order(df2$diff)])
qplot(TVs, mmath, data=df, geom="line", ylab="Average math score") + facet_wrap(~name, ncol=8) + theme_bw()
```